Tidal streams in triaxial systems

### 2015-04-01 ms.tex
phrasing (are we not doing phrasing anymore???)

Figure~\ref{fig:lyap-orbits} shows $l_N$ computed for the two orbits described above. The MLE estimate continues towards zero for the regular orbit (left) but saturates for the chaotic orbit (right). For mildly chaotic or ``sticky'' orbits, it can take integration periods of many thousands of orbits to determine a reliable estimate of the Lyapunov exponent. In comparison, owing to the precision of NAFF, detecting small frequency diffusion rates can be done with only $\sim$100 orbital periods. 
[TODO: axe this figure]
Figure~\ref{fig:fdiff-orbits} shows the three fundamental frequencies computed for 12 consecutive windows of orbit (each section is $\sim$100 orbital periods in length) for Orbit A (left) and Orbit B (right). Dashed horizontal line is the estimate of the uncertainty in the numerical computation of the frequencies from Orbit A. The value of each frequency in the first window is subtracted from each subsequent window. The frequencies only show minor deviations for Orbit A (left panel) whereas there is significant diffusion in Orbit B (right panel).

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant \emph{local} to a given orbit --- that is, these quantities are measures of how infinitesimal deviations will diverge, on average, from some parent orbit of interest. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy, but for a dwarf-galaxy-scale progenitor, the dispersion can be 5--20\%. This raises an important consideration: do finite deviations diverge faster than predicted from local measures of chaos? Does the Lyapunov time or frequency rate predict the timescale over which a finite phase-space volume stays ``coherent''?

It is computationally intractable to run full N-body simulations of stellar system disruption for large grids of orbital initial conditions and thus we take a simplified approach for analyzing these questions. We instead consider small ensembles of particles meant to represent debris disrupted from a single tidal disruption event. For a given set of orbital initial conditions --- the ``parent'' orbit --- we find the position of the nearest pericenter, initialize a small ensemble of test particle orbits around this position, then follow the orbits of all of these test particles for an integration time equal to 256 times the longest fundamental orbital period. The physical scale of the ensemble is set by the tidal radius in position and velocity scale in velocity and are therefore set by the mass scale of the progenitor \citep[e.g.,][]{johnston98, apw14}. If $(\bs{x}_0,\bs{v}_0)$ are the parent orbit initial conditions at pericenter, then $(\delta\bs{x}_i,\delta\bs{v}_i)$ is the deviation vector of the $i$th particle and the magnitude of the offsets are assumed to be Normally distributed away from the parent orbit:
\begin{align}
	\delta\bs{x}_i &\sim \mathcal{N}(0, r_{\rm tide})\\
	\delta\bs{v}_i &\sim \mathcal{N}(0, \sigma_v)\\
	r_{\rm tide} &= \|\bs{x}_0\| \left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} \\
	\sigma_v &= \|\bs{v}_0\|\left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} 
\end{align}
where $M_{\rm enc}(r)$ is the mass enclosed of the host potential within radius $r$, $m$ is the mass scale of the debris, and $\|\cdot \|$ is the Euclidean norm. For this work, we take $m=10^4~\msun$ to represent globular-cluster-like progenitors, and use the spherically-averaged enclosed mass to estimate the above debris scales.

As the particle ensembles evolve in time, we follow a method similar to that of \citet[][hereafter MV96]{merritt96} to quantify the degree of mixing of the ensemble by computing the ``distance'' between the configuration space density of the ensemble at a given time, $\rho(\bs{x},t)$, with that of a distribution that is uniformly distributed over the energy hypersurface, $\rho_{E_0}(\bs{x})$, where $E_0$ is the energy of the parent orbit and thus hypersurface of interest. Our method differs in a few ways from that of MV96. First, MV96 use isoenergy ensembles of particles for following the density because this is the only way to guarantee that, if chaotic, nearly all orbits in the ensemble will lie on chaotic orbits. We instead use ensembles with small but finite spreads in energy; we still expect that most ensemble particle orbits around a parent chaotic orbit will reach a fully mixed state but now the final distribution will have some thickness around this surface. Around a typical regular orbit, each individual orbit will fill its torus ergodically --- a surface with lower dimension than the energy hypersurface --- and thus, a small enough ensemble should never reach the fully mixed state as defined above because it will remain confined to a subspace of lower dimensionality. In practice, only strongly chaotic systems will get close to $\rho_{E_0}(\bs{x})$ in a reasonable amount of time; we expect the rate to depend on how stochastic the parent orbit is (e.g., as measured by the Lyapunov time or frequency diffusion rate) and also on how ``trapped'' the chaotic region is (a small chaotic region wedged between stable tori will take longer to appear chaotic, even if internally the frequency diffusion rate is large). However, the difference in configuration space density between an ensemble undergoing chaotic mixing versus an ensemble around a regular orbit may be detectable over a much shorter timescale. The fully-mixed, normalized, configuration-space density that fills a given energy hypersurface, $E_0$, is easily found by projecting the simple distribution function, $f$, into configuration space:
\begin{align}
	f(\bs{x}, \bs{v}) &= f_0 \, \delta(H(\bs{x}, \bs{v}) - E_0)\\
	\rho_{E_0}(\bs{x}) &= \int \dd^3v \, f(\bs{x}, \bs{v})\\
	&= \int \dd^3v \, f_0 \, \delta(H - E_0)\\
	&= 4\pi \int \dd v \, v^2 \, f_0 \, \delta\left(v^2/2 + \Phi(\bs{x}) - E_0\right)\\
	&= A \sqrt{E_0 - \Phi(\bs{x})}\label{eq:mixeddensity}
\end{align}
where $A$ is a constant determined such that $\int \dd^3\bs{x} \, \rho_{E_0}(\bs{x}) = 1$, and $\delta(\cdot)$ is the Dirac delta function.

>>>>

Figure~\ref{fig:lyap-orbits} shows $l_N$ computed for the two orbits described above. The MLE estimate continues towards zero for the regular orbit (left) but saturates for the chaotic orbit (right). For mildly chaotic or ``sticky'' orbits, it can take integration periods of many thousands of orbits to determine a reliable estimate of the Lyapunov exponent. In comparison, owing to the precision of NAFF, detecting small frequency diffusion rates can be done with only $\sim$100 orbital periods. Figure~\ref{fig:fdiff-orbits} shows the three fundamental frequencies computed for a sliding window of 12 sections of orbit (each section is $\sim$100 orbital periods in length) for Orbit A (left) and Orbit B (right). Dashed horizontal line is the estimate of the uncertainty in the numerical computation of the frequencies from Orbit A. The value of each frequency in the first window is subtracted from each subsequent window. The frequencies only show minor deviations for Orbit A (left panel) whereas there is significant diffusion in Orbit B (right panel).

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant \emph{local} to a given orbit --- that is, these quantities are measures of how infinitesimal deviations will diverge, on average, from some parent orbit of interest. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy, but for a dwarf-galaxy-scale progenitor, the dispersion can be 5--20\%. This raises an important consideration: do finite deviations diverge faster than predicted from local measures of chaos? Does the Lyapunov time or frequency rate predict the timescale over which a finite phase-space volume (e.g., tidal debris) stays ``coherent''?

It is computationally intractable to run full N-body simulations of stellar system disruption for large grids of orbital initial conditions and thus we take a simplified approach for analyzing these questions. We instead consider small ensembles of particles meant to represent debris disrupted from a single tidal disruption event. For a given set of orbital initial conditions --- the ``parent'' orbit --- we find the position of the nearest pericenter, initialize a small ensemble of test particle orbits around this position, then follow the orbits of all test particles for an integration time equal to XXX times the longest fundamental orbital period. The physical scale of the ensemble is set by the tidal radius in position and velocity scale in velocity and are therefore set by the mass scale of the progenitor \citep[e.g.,][]{johnston98, apw14}. If $(\bs{x}_0,\bs{v}_0)$ are the parent orbit initial conditions at pericenter, then $(\delta\bs{x}_i,\delta\bs{v}_i)$ is the deviation vector of the $i$th particle and the magnitude of the offsets are assumed to be Normally distributed away from the parent orbit:
\begin{align}
	\delta\bs{x}_i &\sim \mathcal{N}(0, r_{\rm tide})\\
	\delta\bs{v}_i &\sim \mathcal{N}(0, \sigma_v)\\
	r_{\rm tide} &= \|\bs{x}_0\| \left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} \\
	\sigma_v &= \|\bs{v}_0\|\left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} 
\end{align}
where $M_{\rm enc}(r)$ is the mass enclosed of the host potential within radius $r$, $m$ is the mass scale of the debris, and $\|\cdot \|$ is the Euclidean norm. For this work, we take $m=10^4~\msun$ to represent globular-cluster-like progenitors, and use the spherically-averaged enclosed mass to estimate the above debris scales.

[TODO: wait, actually, they differ in a log -- why doesn't this change things more? I guess this is saying like, mean(log(dens)) ~ log(mean(dens))??? confuzzled]
As the particle ensembles evolve in time, we follow the mean configuration-space density of each to quantify the degree of mixing of the ensembles. This is similar to the method of \citet{merritt96} who measured the evolution of the relative entropy between the configuration-space density of isoenergy ensembles, $\rho(\bs{x},t)$, and the density of an ensemble that is uniformly distributed over the energy hypersurface, $\rho_{E_0}(\bs{x})$. In our case, the

 [but isoenergy so constant...yadda yadda].

to quantify the degree of mixing of the ensemble by computing the ``distance'' between the configuration space density of the ensemble at a given time, $\rho(\bs{x},t)$, with that of a distribution that is uniformly distributed over the energy hypersurface, $\rho_{E_0}(\bs{x})$, where $E_0$ is the energy of the parent orbit and thus hypersurface of interest. Our method differs in a few ways from that of MV96. First, MV96 use isoenergy ensembles of particles for following the density because this is the only way to guarantee that, if chaotic, nearly all orbits in the ensemble will lie on chaotic orbits. We instead use ensembles with small but finite spreads in energy; we still expect that most ensemble particle orbits around a parent chaotic orbit will reach a fully mixed state but now the final distribution will have some thickness around this surface. Around a typical regular orbit, each individual orbit will fill its torus ergodically --- a surface with lower dimension than the energy hypersurface --- and thus, a small enough ensemble should never reach the fully mixed state as defined above because it will remain confined to a subspace of lower dimensionality. In practice, only strongly chaotic systems will get close to $\rho_{E_0}(\bs{x})$ in a reasonable amount of time; we expect the rate to depend on how stochastic the parent orbit is (e.g., as measured by the Lyapunov time or frequency diffusion rate) and also on how ``trapped'' the chaotic region is (a small chaotic region wedged between stable tori will take longer to appear chaotic, even if internally the frequency diffusion rate is large). However, the difference in configuration space density between an ensemble undergoing chaotic mixing versus an ensemble around a regular orbit may be detectable over a much shorter timescale. The fully-mixed, normalized, configuration-space density that fills a given energy hypersurface, $E_0$, is easily found by projecting the simple distribution function, $f$, into configuration space:
\begin{align}
	f(\bs{x}, \bs{v}) &= f_0 \, \delta(H(\bs{x}, \bs{v}) - E_0)\\
	\rho_{E_0}(\bs{x}) &= \int \dd^3v \, f(\bs{x}, \bs{v})\\
	&= \int \dd^3v \, f_0 \, \delta(H - E_0)\\
	&= 4\pi \int \dd v \, v^2 \, f_0 \, \delta\left(v^2/2 + \Phi(\bs{x}) - E_0\right)\\
	&= A \sqrt{E_0 - \Phi(\bs{x})}\label{eq:mixeddensity}
\end{align}
where $A$ is a constant determined such that $\int \dd^3\bs{x} \, \rho_{E_0}(\bs{x}) = 1$, and $\delta(\cdot)$ is the Dirac delta function.
